Constraints on Assembly Bias from Galaxy Clustering

### 2016-06-10 ms.tex
removed obsolete draft comments. removed period from end of title

\title[Clustering Constraints on Assembly Bias]
{
Constraints on Assembly Bias from Galaxy Clustering. 
}

@@

\maketitle
%----------------------------------------------------------------
%%%%%%%%%%%%%%%%%%%%%%%  A B S T R A C T %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ARZ: Important points:
% -supersedes previous analyses even for standard HOD.
% -cannot rule out assembly bias using galaxy clustering.
% -some samples favor assembly bias.
\begin{abstract}
  We constrain the newly-introduced Decorated HOD model using SDSS DR7
  measurements of projected galaxy clustering, $\wprp,$ and number
  density, $n_{\rm d},$ made from r-band luminosity-threshold
  samples. The Decorated HOD is a model for the galaxy--halo
  connection that augments the traditional Halo Occupation
  Distribution (HOD) by allowing for the possibility of {\em assembly
    bias}: galaxy luminosity may be correlated with dark matter halo
  properties besides virial mass $\mvir$ alone. We demonstrate that it
  is not possible to rule out assembly bias using DR7 measurements of
  $\wprp$ and $n_{\rm d}.$ Moreover, galaxy samples $\magr<-20, -20.5$
  favor strong levels of central galaxy assembly bias:
  high-concentration halos are more likely to host a central galaxy
  relative to low-concentration halos of the same $\mvir.$ We rule out
  zero assembly bias with high significance for these
  samples. Satellite assembly bias only becomes significant for the
  faintest sample we study, $\magr<-19.$ We find no evidence for
  assembly bias in the $\magr<-21$ sample. Because the Decorated HOD
  subsumes the traditional HOD, and additionally because of errors
  made in the original analysis of these samples, our results differ
  from and supersede existing HOD constraints from low-redshift
  clustering measurements.
\end{abstract}

@@

For more than a decade, halo occupation modeling has been used to
interpret large-scale structure measurements and exploit these
measurements to constrain galaxy formation models and cosmology
\citep[e.g.,][]{yang03,tinker05,zehavi05a,
  porciani06,vdBosch07,Zheng07,conroy_wechsler09,yang09b,zehavi_etal11,guo_etal11b,
  wake_etal11,yang11a,yang12,leauthaud_etal12,rod_puebla12,tinker_etal13,cacciato_etal13,
  more_etal13,guo_etal14,zu_mandelbaum15b}. The key assumptions
underlying halo occupation modeling are: (1) all galaxies reside in
dark matter halos that are biased tracers of the density field; and
(2) galaxies occupy halos as a function of halo masses only. It is now
well known that halo bias depends upon halo properties other than mass
\citep[e.g.][]{gao_etal05,wechsler06,gao_white07,zentner07,dalal_etal08,lacerna11},
an effect called halo assembly bias. If galaxies occupy halos as a
function of properties other than halo mass, then standard halo
occupation methods will be subject to a systematic error due to galaxy
assembly bias. Several of us have previously shown that this error can
be significant in an analysis of galaxy clustering and can bias
inferences about many aspects of galaxy evolution
\citep{zentner_etal14}. Consequently, we have developed halo
occupation models that enable galaxies to occupy halos in a manner
that depends upon several halo properties \citep{hearin_etal16}.  In
this paper, we revisit the interpretation of luminosity-dependent
galaxy clustering in the Sloan Digital Sky Survey (SDSS) Data Release
7 (DR7) data, analyzed previously by \citet{zehavi_etal11}, in the
context both standard halo occupation models and these new models.

The ultimate goal of halo occupation modeling is to characterize the
galaxy-dark matter connection by constraining the probability
distribution $P(X_{\rm g}|Y_{\rm h})$. Here $X_{\rm g} = \{x_1, x_2,
  ...x_n\}$ is a set of $n$ galaxy properties, and $Y_{\rm h} = \{y_1,
  y_2,...,y_m\}$ a set of $m$ halo properties. Accurate knowledge of
this probability distribution yields invaluable insight regarding the
processes of galaxy formation, and allows one to make accurate
predictions for the clustering of galaxies as function of any property
$x_i$. Most halo occupation studies to date have restricted themselves
to only one halo property, namely halo mass.  This includes Halo
Occupation Distribution (HOD) models, which aim to constrain
$P(N|\mvir)$, the probability that a halo of mass $M$ contains $N$
galaxies of some specified properties
\citep[e.g.,][]{kauffmann_etal97,jing98,benson01,berlind02}, and
Conditional Luminosity Function (CLF) models, which aim to constrain
$\Phi(L|M)$, the luminosity (or stellar mass) function of galaxies in
haloes of mass $M$ \citep[e.g.,][]{yang03,vdBosch03a,cooray06}. The
HOD and CLF models are mainly used in combination with the halo model,
which describes the non-linear matter field in terms of halo building
blocks \citep[e.g.,][]{seljak00,cooray02}, to predict the clustering
of galaxies as function of luminosity or stellar mass.  Although
entirely analytical, and therefore fast, the downside of this method
is that it is extremely challenging to construct a model that is
accurate at the few percent level. The main issues are halo exclusion
\citep[see][]{smith_etal07,vdBosch13}, scale dependence of the halo
bias \citep[see][]{tinker05}, and the halo assembly bias mentioned
above.

These issues are trivially avoided by directly populating dark matter
haloes in numerical simulations with galaxies. This can be done using
either a parameterized model for the halo occupation statistics
\citep[e.g.,][]{Yang_etal04,moster10,behroozi10,behroozi13} or by
rank-order matching galaxy properties (i.e., luminosity or stellar
mass) to halo properties (i.e., halo mass). The latter method is known
as subhalo abundance matching (SHAM).  Although many SHAM studies use
halo mass as the halo property of choice
\citep[e.g.,][]{vale_ostriker04,vale_ostriker06,conroy07,conroy_wechsler09},
a number of studies have used alternative halo properties, such as the
maximum circular velocity, $\vmax$, or its peak value over the halo's
history, $V_{\rm peak}$
\citep[e.g.][]{tasitsiomi_etal04,conroy06,marin_etal08,reddick_etal13}

Since, at fixed halo mass, $\vmax$ is correlated with halo
concentration, and thus with halo assembly time, the resulting
occupation models have assembly bias build in
\citep[see][]{zentner_etal14}; i.e., at fixed halo mass, galaxy
properties will be correlated with halo assembly
time. \citet{lehmann_etal15} took this one step further, and explored
SHAM models in which the halo property varies between $M$ and $\vmax$
via a continuously-valued parameter. Using SDSS clustering
measurements to constrain their model, they conclude that at fixed
halo mass galaxy luminosity reveals a significant dependence on halo
concentration. Since the latter is strongly correlated with halo
formation time \citep[e.g.,][]{wechsler02,zhao_etal09,ludlow_etal13},
their conclusions argue for non-zero assembly bias in the galaxy
population.

Additional support for the presence of assembly bias has come from a
variety of studies that have shown, or suggested, that the large-scale
environment (i.e., clustering strength) of dark matter haloes {\it of
  fixed mass} is correlated with the star formation rate of their
central
galaxies\citep[][]{yang_etal06a,blanton_berlind07,wang_etal08,tinker_etal12,wang_etal13,HW13a,hearin_etal13b,hearin_etal14,watson_etal14}.
However, \citet{lin_mandelbaum_etal15} claim that most of these
differences in clustering strength can be explained as due to
differences in halo mass or contamination by satellite galaxies, and
therefore argue that there is no significant evidence for assembly
bias in galaxy-scale halos \citep[see also][]{paranjape15}.  On the
other hand, \citet{miyatake_etal16} and \citet{more_etal16} recently
presented strong evidence for the presence of assembly bias in massive
clusters; in particular, they find that clusters with a more
concentrated distribution of satellite galaxies are less strongly
clustered than their equal-mass counterparts whose satellite galaxies
are less centrally concentrated. To summarize, while there is strong
evidence for the presence of assembly bias on the scale of clusters,
there is not concensus yet as to whether assembly bias plays also a
role in less massive haloes.

Our work is important for several reasons. Our work is a re-analysis
of the SDSS DR7 data that overcomes shortcomings of previous
analyses. In particular, we use direct population of galaxies in a
cosmological simulation so that delicate issues present in analytic
modeling, such as scale-dependent halo bias, are treated exactly.  The
simulation we use is based on the latest Planck cosmological
parameters, updating previous work. Furthermore, many important
differences between this work and the previous analysis by
\citet{zehavi_etal11} result from the fact that the Monte Carlo Markov
Chains used by Zehavi et al. were too small and had therefore not
converged (Z. Zheng \& I. Zehavi, private communication).  Therefore,
our analysis {\em supersedes previous analyses even in the case of
  standard halo occupation models}.

Furthermore, our work demonstrates explicitly that significant
assembly bias in $M_r$-selected samples from SDSS DR7 cannot be ruled
out based on a standard analysis of galaxy clustering only. In fact,
in agreement with \citet{lehmann_etal15}, we find that several samples
{\em favor galaxy assembly bias to a degree that is statistically
  significant}. As demonstrated by \citet{zentner_etal14}, this
conclusion could have important consequences for the interpretation of
both extant and forthcoming data.

Our paper is organized as follows. In Section~\ref{section:methods},
we discuss our implementation of halo occupation models and our
assumptions in our parameter inference analysis. We present results
for both standard halo occupation analysis and analysis in the context
of models with galaxy assembly bias in
Section~\ref{section:results}. We summarize our results and draw
conclusions in Section~\ref{section:conclusions}.

>>>>

\title[Clustering Constraints on Assembly Bias]
{
Constraints on Assembly Bias from Galaxy Clustering
}

@@

\maketitle
%----------------------------------------------------------------
%%%%%%%%%%%%%%%%%%%%%%%  A B S T R A C T %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
  We constrain the newly-introduced Decorated HOD model using SDSS DR7
  measurements of projected galaxy clustering, $\wprp,$ and number
  density, $n_{\rm d},$ made from r-band luminosity-threshold
  samples. The Decorated HOD is a model for the galaxy--halo
  connection that augments the traditional Halo Occupation
  Distribution (HOD) by allowing for the possibility of {\em assembly
    bias}: galaxy luminosity may be correlated with dark matter halo
  properties besides virial mass $\mvir$ alone. We demonstrate that it
  is not possible to rule out assembly bias using DR7 measurements of
  $\wprp$ and $n_{\rm d}.$ Moreover, galaxy samples $\magr<-20, -20.5$
  favor strong levels of central galaxy assembly bias:
  high-concentration halos are more likely to host a central galaxy
  relative to low-concentration halos of the same $\mvir.$ We rule out
  zero assembly bias with high significance for these
  samples. Satellite assembly bias only becomes significant for the
  faintest sample we study, $\magr<-19.$ We find no evidence for
  assembly bias in the $\magr<-21$ sample. Because the Decorated HOD
  subsumes the traditional HOD, and additionally because of errors
  made in the original analysis of these samples, our results differ
  from and supersede existing HOD constraints from low-redshift
  clustering measurements.
\end{abstract}

@@

For more than a decade, halo occupation modeling has been used to
interpret large-scale structure measurements and exploit these
measurements to constrain galaxy formation models and cosmology
\citep[e.g.,][]{yang03,tinker05,zehavi05a,
  porciani06,vdBosch07,Zheng07,conroy_wechsler09,yang09b,zehavi_etal11,guo_etal11b,
  wake_etal11,yang11a,yang12,leauthaud_etal12,rod_puebla12,tinker_etal13,cacciato_etal13,
  more_etal13,guo_etal14,zu_mandelbaum15b}. The key assumptions
underlying halo occupation modeling are: (1) all galaxies reside in
dark matter halos that are biased tracers of the density field; and
(2) galaxies occupy halos as a function of halo mass $\mvir$ only.

It is now well known that halo bias depends upon halo properties other than $\mvir$
\citep[e.g.][]{gao_etal05,wechsler06,gao_white07,zentner07,dalal_etal08,lacerna11},
an effect called halo assembly bias. If galaxies occupy halos as a
function of halo properties other than $\mvir,$ then standard halo
occupation methods will be subject to a systematic error due to galaxy
assembly bias. 
Several of us have previously shown that this error can
be significant in an analysis of galaxy clustering and can bias
inferences about many aspects of galaxy evolution
\citep{zentner_etal14}.

There is increasing observational evidence that galaxy assembly bias is 
present in the real Universe. 
Using an approach based on subhalo abundance matching, 
\citet{lehmann_etal15} showed that the clustering of low-redshift 
galaxies favors models in which stellar mass depends upon a 
combination of $\mvir$ and halo concentration.
\citet{miyatake_etal16} and \citet{more_etal16} have 
presented evidence for the presence of assembly bias in massive
clusters; in particular, they find that clusters with a more
concentrated distribution of satellite galaxies exhibit a weaker large-scale bias 
relative to clusters of the same mass with satellite galaxies that
are less centrally concentrated. 
Additional support for the presence of assembly bias has come from a
variety of studies that have shown, or suggested, that the large-scale
environment of dark matter halos of fixed mass 
is correlated with the star formation rate of their central galaxies 
\citep{yang_etal06a, blanton_berlind07,wang_etal08,wang_etal13,hearin_etal14}.

Despite this growing evidence, the empirical modeling community 
has not yet reached a consensus on the question of galaxy assembly bias. 
For example, \citet{lin_mandelbaum_etal15} claim that 
some of the above evidence for assembly bias can be explained by
differences in halo mass between the galaxy samples under consideration.
\citet{paranjape15} urges caution that contamination by satellite galaxies 
can masquerade as a false signal of assembly bias. 
\citet{tinker08b} argue that measurements of void statistics place strong 
bounds on the possible strength of the signal.

In much of the literature on this topic, 
the character of the supporting evidence
suffers from a severe shortcoming. 
Until very recently, the only models that have been fit to 
observational clustering measurements have been traditional models 
in which assembly bias is assumed from the outset to be zero. 
Models that include assembly bias such as semi-analytic models 
are typically too computationally expensive 
for anything besides a ``one-off" comparison \citep[e.g.,][]{croton_etal07}, prohibiting 
any conclusive statement to be made about the strength of the signal.

Motivated by this shortcoming, we have developed a new class of 
empirical models that enable galaxies to occupy halos in a manner
that has {\em simultaneous} dependence 
upon multiple halo properties \citep{hearin_etal16}, 
{\em including continuously variable levels of assembly bias}.
Crucially, our implementation is of sufficient computational efficiency 
to permit a proper likelihood analysis of the model parameter space. 
Armed with this new methodology, in this paper 
we revisit the interpretation of luminosity-dependent
galaxy clustering in the Sloan Digital Sky Survey (SDSS) Data Release
7 (DR7) data, analyzed previously by \citet{zehavi_etal11}, in the
context both standard halo occupation models and these new models.

Our work is a re-analysis of the SDSS DR7 data that 
overcomes numerous shortcomings of previous
analyses. In particular, we use direct population of galaxies in a
cosmological simulation so that delicate issues present in analytic
modeling, such as scale-dependent halo bias, are treated exactly.  The
simulation we use is based on the latest Planck cosmological
parameters \citep{planck13}, updating previous work. 
Furthermore, many important
differences between this work and the previous analysis by
\citet{zehavi_etal11} result from the fact that the Monte Carlo Markov
Chains used in \citet{zehavi_etal11} were too small and had not
converged (Z. Zheng \& I. Zehavi, private communication).  Therefore,
{\em our analysis supersedes previous work even in the case of
standard halo occupation models}.

Most importantly, our work demonstrates explicitly that significant
assembly bias in $\magr$-selected samples from SDSS DR7 cannot be ruled
out based on a standard analysis of galaxy clustering only. In fact,
in agreement with \citet{lehmann_etal15}, we find that several samples
{\em favor galaxy assembly bias to a degree that is statistically
  significant}. As demonstrated by \citet{zentner_etal14}, this
conclusion has important consequences for the interpretation of
both extant and forthcoming data.

Our paper is organized as follows. In Section~\ref{section:methods},
we discuss our implementation of halo occupation models and the
parameter inference methodology. We present results
from both traditional and assembly-biased halo occupation analyses 
in Section~\ref{section:results}. We summarize our results and draw
conclusions in Section~\ref{section:conclusions}.
